\thesisTitle

### 2019-10-29 proposal/paper.tex
minor corrections

		Recently the class of \textit{convolutional graph neural networks} (ConvGNNs) gained traction.
		It generalizes the convolution operator that is used by \textit{convolutional neural networks} (CNNs) to graphs.
		There are two main variants of ConvGNNs:
		\begin{enumerate}[label=\textbf{\alph*)}] % chktex 9 chktex 10
			\item \textbf{Spatial variant:}
				Standard CNNs use a convolution operator that is typically applied to a grid graph of pixels with added diagonal edges.
				Spatial ConvGNNs~\cite{Micheli2009} directly generalize this idea by defining a more flexible convolution that works with arbitrary graph structures.
				Spatial graph convolution aggregates the direct neighborhood of a vertex $v_i$, where $v_i$ is typically described by a feature vector $x_i \in \mathbb{R}^d$.
				This process returns a new aggregate feature vector $x'_i$.
				By stacking multiple convolution layers, features from indirect neighbors become part of the aggregate.
				Analogous to how CNNs learn kernel matrices, spatial ConvGNNs learn vertex neighborhood feature aggregation functions.
				This approach shares similarities with the previously mentioned message-passing architecture~\cite{Gori2005} with the major different being that the number of message-passing, i.e.\ convolution, steps is fixed.
			\item \textbf{Spectral variant:}
				Motivated by the theoretical foundation provided by spectral graph theory~\cite{Shuman2013}, spectral ConvGNNs~\cite{Bruna2013} learn a filter function $\hat{g}$ that is applied in the Fourier domain of a graph.
				Based on the convolutional theorem, a convolution operator can not only be expressed in terms of the neighborhood of vertices but also as a filter on the eigenvectors of a graph's Laplacian.
				Formally this can be expressed as
				\begin{align}
					g *_G \bm{x} = U \hat{g}(\Lambda) U^{\top} \bm{x}\label{eq:related-work:conv-filter}
				\end{align}
				where $\bm{x} \in \mathbb{R}^{n \times k}$ is the matrix of all vertex features and $U \Lambda U^{\top}$ is the eigendecomposition of the graph Laplacian $L_G$\footnote{
					To see the connection to the convolutional theorem, note that $\mathcal{F} = U$ and $\mathcal{F}^{-1} = U^{\top}$ in \cref{eq:related-work:conv-filter} with $\mathcal{F}$ denoting the graph Fourier transform.
				}.
				In the previously described spatial ConvGNN variant, $g$ is learned directly in form of the feature aggregation function with a neighborhood locality constraint.
				In the spectral variant however the Fourier-transformed filter $\hat{g}$ is learned.
				Intuitively this means that vertex features $x_i$ are not aggregated with the features of their their direct neighborhood but with the features of vertices with which they share certain structural similarities.
				This allows spectral ConvGNNs to incorporate the global graph structure at the cost of having to perform the computationally expensive decomposition of every input graph's Laplacian.

@@

		The just described ConvGNN approaches generally assume that an input consists of an undirected graph with positive real edge weights.
		Those approaches are not directly applicable in scenarios with multiple edge types, i.e.\ multi-relational scenarios.
		One class of spatial multi-relational GNN methods was proposed by \citet{Battaglia2018}.
		They describe a general framework for the construction of GNNs; for simplicity only the class message-passing variants is mentioned here.
		There a shared function $\phi^e$ is used to describe how a vertex feature vector $v_i$ is transformed as it passes through an edge with features $e_{ij} \in \mathbb{R}^p$.
		In a uni-relational spatial GNN with $e_{ij} \in \mathbb{R}^{1}$, typically $\phi^e(e_{ij}, v_i) = e_k \cdot v_i$ is used.
		In the multi-relational setting, $\phi^e$ can instead be learned via a standard \textit{multilayer perceptron} (MLP).
		An alternative spectral multi-relational ConvGNN approach was proposed by \citet{Gong2018}.
		Their so-called \textit{edge enhanced graph neural network} (EGNN) essentially models each relation as a separate graph with its own Laplacian.
		A convolutional filter $g$ is shared across all relations and used to convolve the input signal separately via each relation's Laplacian.
		To share vertex features across relations, the separately convolved signals of all relations are then concatenated.
		Those concatenated vertex features are used as the input signal of the next layer.

@@

Intuitively the difference between LTA and current graph methods can be formulated as \textit{first evaluate, then aggregate} versus \textit{first aggregate, then evaluate}, see \cref{fig:gnnvslta}.
This general difference is however not a strict separator of LTA and current graph methods.
As described by \citet{Pope2018} for example, relevant subgraphs with a high influence on the output of a ConvGNN can be identified and interpreted similarly to LTA constituents.
The line between both approaches is therefore blurry.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\linewidth]{GNNvsLTA.pdf}
	\caption{Intuition for the difference between ConvGNNs and LTA methods.}\label{fig:gnnvslta}
\end{figure}

To address this, the thesis should try to answer three central questions:
\begin{enumerate}
	\item \textit{What are the essential characteristics of a Learning to Aggregate method?}
		Those characteristics should formally capture the differences and similarities between LTA and convolutional or kernel based graph methods.
		Currently there is no comprehensive formulation of the relation between both fields of research;
		this should be addressed by the first question.
	\item \textit{How could a Learning to Aggregate method, that has the previously defined characteristics, look like for graph inputs?}
		Currently LTA has only been applied to sets of constituents and not yet to graph structured data;
		this should be addressed by the second question.
	\item \textit{What are the assumptions of graph LTA methods and under which circumstances are they applicable?}
		This aims to clarify in which kinds of scenarios the use of LTA is justified and expected to improve results compared to more general methods.
\end{enumerate}

@@

The resulting general definition of LTA methods should then be used for a comparison with a representative selection of existing graph methods (see \cref{sec:related-work:gcr}).
It should be stated how both fields of research are related and under which circumstances a graph method can be interpreted as an LTA method.
When formulating a graph method from the LTA perspective, it must be explicitly stated how the constituents and the aggregation function looks like.
Depending on the graph method a constituent could consist of a single vertex but could also represent a certain subgraph or characteristic graph pattern.

@@

Finally the proposed graph LTA method should be evaluated.
For this synthetically generated data or some of the datasets listed in \cref{sec:related-work:tox} could be used.
If time permits, a comparison between the developed graph LTA method and other graph methods should be conducted to empirically demonstrate the potential advantages and disadvantages of the LTA assumptions.

>>>>

		Recently the class of \textit{convolutional graph neural networks} (ConvGNNs) gained traction.
		It generalizes the convolution operator that is used by \textit{convolutional neural networks} (CNNs) to graphs.
		There are two main variants of ConvGNNs:
		\begin{enumerate}[label=\textbf{\alph*)}] % chktex 9 chktex 10
			\item \textbf{Spatial variant:}
				Standard CNNs use a convolution operator that is typically applied to a grid graph of pixels with added diagonal edges.
				Spatial ConvGNNs~\cite{Micheli2009} directly generalize this idea by defining a more flexible convolution that works with arbitrary graph structures.
				Spatial graph convolution aggregates the direct neighborhood of a vertex $v_i$, where $v_i$ is typically described by a feature vector $x_i \in \mathbb{R}^d$.
				This process returns a new aggregate feature vector $x'_i$.
				By stacking multiple convolution layers, features from indirect neighbors become part of the aggregate.
				Analogous to how CNNs learn kernel matrices, spatial ConvGNNs learn vertex neighborhood feature aggregation functions.
				This approach shares similarities with the previously mentioned message-passing architecture~\cite{Gori2005} with the major different being that the number of message-passing, i.e.\ convolution, steps is fixed.
			\item \textbf{Spectral variant:}
				Motivated by the theoretical foundation provided by spectral graph theory~\cite{Shuman2013}, spectral ConvGNNs~\cite{Bruna2013} learn a filter function $\hat{g}$ that is applied in the Fourier domain of a graph.
				Based on the convolutional theorem, a convolution operator can not only be expressed in terms of the neighborhood of vertices but also as a filter on the eigenvectors of a graph's Laplacian.
				Formally this can be expressed as
				\begin{align}
					g *_G \bm{x} = U \hat{g}(\Lambda) U^{\top} \bm{x}\label{eq:related-work:conv-filter}
				\end{align}
				where $\bm{x} \in \mathbb{R}^{n \times k}$ is the matrix of all vertex features and $U \Lambda U^{\top}$ is the eigendecomposition of the graph Laplacian $L_G$\footnote{
					To see the connection to the convolutional theorem, note that $\mathcal{F} = U$ and $\mathcal{F}^{-1} = U^{\top}$ in \cref{eq:related-work:conv-filter} with $\mathcal{F}$ denoting the graph Fourier transform.
				}.
				In the previously described spatial ConvGNN variant, $g$ is learned directly in form of the feature aggregation function with a neighborhood locality constraint.
				In the spectral variant however, the Fourier-transformed filter $\hat{g}$ is learned.
				Intuitively this means that vertex features $x_i$ are not aggregated with the features of their their direct neighborhood but with the features of vertices with which they share certain structural similarities.
				This allows spectral ConvGNNs to incorporate the global graph structure at the cost of having to perform the computationally expensive eigendecomposition of every input graph's Laplacian.

@@

		The just described ConvGNN approaches generally assume that an input consists of an undirected graph with positive real edge weights.
		Those approaches are not directly applicable in scenarios with multiple edge types, i.e.\ multi-relational scenarios.
		One class of spatial multi-relational GNN methods was proposed by \citet{Battaglia2018}.
		They describe a general framework for the construction of GNNs; for simplicity only the class of message-passing variants is mentioned here.
		There a shared function $\phi^e$ is used to describe how a vertex feature vector $v_i$ is transformed as it passes through an edge with features $e_{ij} \in \mathbb{R}^p$.
		In a uni-relational spatial GNN with $e_{ij} \in \mathbb{R}^{1}$, typically $\phi^e(e_{ij}, v_i) = e_k \cdot v_i$ is used.
		In the multi-relational setting, $\phi^e$ can instead be learned via a standard \textit{multilayer perceptron} (MLP).
		An alternative spectral multi-relational ConvGNN approach was proposed by \citet{Gong2018}.
		Their so-called \textit{edge enhanced graph neural network} (EGNN) essentially models each relation as a separate graph with its own Laplacian.
		A convolutional filter $g$ is shared across all relations and used to convolve the input signal separately via each relation's Laplacian.
		To share vertex features across relations, the separately convolved signals of all relations are then concatenated.
		Those concatenated vertex features are used as the input signal of the next layer.

@@

Intuitively the difference between LTA and current graph methods can be formulated as \textit{first evaluate, then aggregate} versus \textit{first aggregate, then evaluate} (see \cref{fig:gnnvslta}).
This general difference is however not a strict separator of LTA and current graph methods.
As described by \citet{Pope2018} for example, relevant subgraphs with a high influence on the output of a ConvGNN can be identified and interpreted similarly to LTA constituents.
The line between both approaches is therefore blurry.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\linewidth]{GNNvsLTA.pdf}
	\caption{Intuition for the difference between ConvGNNs and LTA methods.}\label{fig:gnnvslta}
\end{figure}

To address this, the thesis should try to answer three central questions:
\begin{enumerate}
	\item \textit{What are the essential characteristics of a Learning to Aggregate method?}
		Those characteristics should formally capture the differences and similarities between LTA and convolutional or kernel-based graph methods.
		Currently there is no comprehensive formulation of the relation between both fields of research;
		this should be addressed by the first question.
	\item \textit{How could a Learning to Aggregate method, that has the previously defined characteristics, look like for graph inputs?}
		Currently LTA has only been applied to multisets of constituents and not yet to graph structured data;
		this should be addressed by the second question.
	\item \textit{What are the assumptions of graph LTA methods and under which circumstances are they applicable?}
		This aims to clarify in which kinds of scenarios the use of LTA is justified and expected to improve results compared to more general methods.
\end{enumerate}

@@

The resulting general definition of LTA methods should then be used for a comparison with a representative selection of existing graph methods (see \cref{sec:related-work:gcr}).
It should be stated how both fields of research are related and under which circumstances a graph method can also be interpreted as an LTA method.
When formulating a graph method from the LTA perspective, it must be explicitly stated how the constituents and the aggregation function look like.
Depending on the graph method a constituent could consist of a single vertex but could also represent a certain subgraph or characteristic graph pattern.

@@

Finally the proposed graph LTA method should be evaluated.
Synthetically generated data or some of the datasets listed in \cref{sec:related-work:tox} could be used for this.
Alternatively the sentiment analysis problem from \textit{natural language processing} (NLP) could be used by interpreting the dependency parse graph of a text as the input and the text's sentiment as its aggregate score.
If time permits, a comparison between the developed graph LTA method and other graph methods should be conducted to demonstrate the potential advantages and disadvantages of the LTA assumptions.
